---
title: "R Environment and Syntax"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
---

```{r installation, echo = FALSE, eval = FALSE}
install.packages("xaringan")
# install.packages("devtools")
devtools::install_github("gadenbuie/xaringanthemer")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(xaringan)
library(xaringanthemer)

write_xaringan_theme() #explore diff theme settings by pressing the 'tab' button
#or just just create a Ninja Themed Presentation from the New R Markdown Document menu in RStudio

mono_light(
  base_color = "#1c5253",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Droid Mono")
)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

